---
title: "Hands-on Exercise 2A"
---

# Overview

Spatial Point Pattern Analysis (SPPA) is the evaluation of the pattern or distribution of a set of points on a surface. The points may represent:

- events such as crimes, traffic accidents, or disease onsets

- business services (e.g., coffee shops and fast-food outlets) or facilities such as childcare centres and eldercare centres.

First-order Spatial Point Pattern Analysis (1st-SPPA) focuses on understanding the intensity or density of points across a study area. It examines how the distribution of points varies over space, essentially identifying trends or patterns in point density. This type of analysis deals with the individual locations of points and their distribution, without considering interactions between them.

# Getting Started

## Importing of Data

The following R packages will be used : - sf : to import manage vector based geospatial data in R - spatstat. SPPA analyses can be conducted and derive kernal desity estimation (KDE) layers - terra : helps to create, read, manipulate and write raster and vector data, faster for larger datasets. - tmap : for cartographic quality static point pattern maps/interactive maps - rvest : to scrap (harvest) data from webpages

```{r}
pacman::p_load(sf, terra, spatstat,tmap, rvest, tidyverse)
```

Now we load in the files!

```{r}
library(sf)
library(tidyverse)

mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson")
childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson") 


```

## Data Preparations

Similar preparations as exercise 1B

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

```

```{r}
childcare_sf <- st_transform(childcare_sf, 3414)

mpsz_cl <-st_transform(mpsz, 3414)
```



# Plotting of dataset 

First we plot out the map

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(childcare_sf), 
     add = TRUE)
```

Alternatively, we can also prepare an interactive point symbol map by using the code chunk below.

```{r}
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
```

::: Warning
Always remember to switch back to plot mode after the interactive map. This is because, each interactive mode will consume a connection. You should also avoid displaying ecessive numbers of interactive maps (i.e. not more than 10) in one RMarkdown document when publish on Netlify.
:::

```{r}
tmap_mode('plot')
```


# Geospatial Data wrangling 
spatstat relies on its own specific data structures like ppp (planar point pattern) for point data and owin for observation windows. 

Now we will convert sd objects into statstat ppp and owin. 

## Convert from sf to ppp 

spatstat requires the point event data in ppp object form. The code chunk below uses [as.ppp()] of spatstat package to convert childcare_sf to ppp format.

we will verify class using class() and summary() to look at the quick stats of the newly converted ppp

```{r}
childcare_ppp <- as.ppp(childcare_sf)

class(childcare_ppp)

summary(childcare_ppp)

```

## Creating owin object

it is a good practice to confine the analysis with a geographical area like Singapore boundary. In spatstat, an object called owin is specially designed to represent this polygonal region.

```{r}
sg_owin <- as.owin(mpsz_cl)
class(sg_owin)
plot(sg_owin)

```

## Combining point event and owin 
In this last step of geospatial data wrangling, we will extract childcare events that are located within Singapore and The output object combined both the point and polygon feature in one ppp object class as shown below.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]

childcareSG_ppp

```

# Nearest Neighbour Analysis 

Nearest Neighbor Analysis (NNA) is a spatial statistics method that calculates the average distance between each point and its closest neighbor to determine if a pattern of points is clustered, dispersed, or randomly distributed.

Clark-Evans test is a specific statistical method used within NNA to quantify whether a point pattern is clustered, random, or uniformly spaced, using the Clark-Evans aggregation index (R) to describe this pattern. NNA provides a numerical value that describes the degree of clustering or regularity, and the Clark-Evans test calculates a specific index (R) for this purpose


We will perform the Clark-Evans test of aggregation for a spatial point pattern by using clarkevans.test() of spatstat.explore package.

The test hypotheses are:

- Ho = The distribution of childcare services are randomly distributed.

- H1= The distribution of childcare services are not randomly distributed.

The 95% confident interval will be used.

## Clark-Evans test without CSR
```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"))
```
::: callout-note
The Clark-Evans test statistic R = 0.49444 is significantly less than 1 and a P value < 2.2e-16

Therefore we can reject the Null hypothesis that The distribution of childcare services are randomly distributed without the CSR. Since p< 0.05. 

We have proven that the distribution of childcare services in Singapore are highly clustered, where the nearest-neighbor distance between childcare centers is roughly half of what we would expect if they were randomly distributed (R value of 0.49)

:::

##Clark-Evans test WITH CSR 

The argument method = “MonteCarlo” is used here. 

The p-value for the test is computed by comparing the observed value of R to the results obtained from nsim (i.e. 39, 99, 999) simulated realisations of Complete Spatial Randomness conditional on the observed number of points.

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)


```

::: callout-note
The Clark-Evans test statistic R = 0.4944 is significantly less than 1 and a P value =0.01

Therefore we can reject the Null hypothesis that The distribution of childcare services are randomly distributed without the CSR. Since p< 0.05. 

We have proven that the distribution of childcare services in Singapore are highly clustered, where the nearest-neighbor distance between childcare centers is roughly half of what we would expect if they were randomly distributed (R value of 0.49)

:::

#Kernal Density Estimation method

Kernel Density Estimation (KDE) is a valuable tool for visualising and analyzing first-order spatial point patterns. It is widely considered a method within Exploratory Spatial Data Analysis (ESDA) because it’s used to visualize and understand spatial data patterns by transforms discrete point data (like locations of childcare service, crime incidents or disease cases) into continuous density surfaces that reveal clusters and variations in event occurrences, without making prior assumptions about data distribution. It helps to begin understanding data distribution, identify hotspots, and explore relationships between spatial variables before performing more rigorous analysis.

## Automatic bandwidth selection method

The following configurations of density() of spatstat is used : 
- bw.diggle() automatic bandwidth selection method. Other recommended methods are bw.CvL(), bw.scott() or bw.ppl().

- The smoothing kernel used is gaussian, which is the default. Other smoothing methods are: “epanechnikov”, “quartic” or “disc”.

- The intensity estimate is corrected for edge effect bias by using method described by Jones (1993) and Diggle (2010, equation 18.9). The default is FALSE.

```{r}
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 

```

now we plot to display the kernal density derived 
```{r}
plot(kde_SG_diggle)

summary(kde_SG_diggle)
```

We can retrive the bandwidth used to compete the kde layer by using the code chunk below :
```{r}
bw <- bw.diggle(childcareSG_ppp)
bw

```

##Rescaling KDE values 

To rescale the KDE values, rescale.ppp() is used to covert the unit of measurement from meter to kilometer

```{r}
childcareSG_ppp_km <- rescale.ppp(
  childcareSG_ppp, 1000, "km")

```

Now, we can re-run density() using the resale data set and plot the output kde map.

```{r}
kde_childcareSG_km <- density(childcareSG_ppp_km,
                              sigma=bw.diggle,
                              edge=TRUE,
                              kernel="gaussian")

plot(kde_childcareSG_km)
```

::: callout-note
Notice that output image looks identical to the earlier version, the only changes in the data values (legend).
:::

## Different automatic bandwidth methods

Besides bw.diggle(), there are three other spatstat functions can be used to determine the bandwidth, they are: bw.CvL(), bw.scott(), and bw.ppl().

### bw.CvL
```{r}
bw.CvL(childcareSG_ppp_km)
```


### bw.scott
```{r}
bw.scott(childcareSG_ppp_km)
```


### bw.ppl
```{r}
bw.ppl(childcareSG_ppp_km)
```

### bw.diggle 

```{r}
bw.diggle(childcareSG_ppp_km)
```

The code below can be used to compare the output of bw.diggle and bw.ppl

```{r}
kde_childcareSG.ppl <- density(childcareSG_ppp_km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")


```

## Different kernal methods 

By default, the kernel method used in density.ppp() is gaussian. But there are three other options, namely: Epanechnikov, Quartic and Dics.

The code chunk below will be used to compute three more kernel density estimations by using these three kernel function.

```{r}
par(mfrow=c(2,2))
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")


```

# Fixed and Adaptive KDE 
## Computing KDE by fixed bandwidth

By defining a bandwidth of 600m, let us compute a KDE layer. 

```{r}
kde_childcareSG_fb <- density(childcareSG_ppp_km,
                              sigma=0.6, 
                              edge=TRUE,
                              kernel="gaussian")
plot(kde_childcareSG_fb)

```

::: callout-note 
the sigma value used is 0.6 since the unit of measurement of childcareSG_ppp_km object is in kilometer, hence the 600m is 0.6km.
:::

## Adaptive Bandwidth

Fixed bandwidth method is very sensitive to highly skew distribution of spatial point patterns over geographical units for example urban versus rural. One way to overcome this problem is by using adaptive bandwidth instead.

`In this section, you will learn how to derive adaptive kernel density estimation by using density.adaptive() of spatstat.

```{r}
kde_childcareSG_ab <- adaptive.density(
  childcareSG_ppp_km, 
  method="kernel")
plot(kde_childcareSG_ab)
```

The fixed and adaptive KDEs can be compared using this: 
```{r}
par(mfrow=c(1,2))
plot(kde_childcareSG_fb, main = "Fixed bandwidth")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
```


# Plotting cartographic quality KDE map

## Converting gridded output into raster

```{r}
kde_childcareSG_bw_terra <- rast(kde_childcareSG_km)

class(kde_childcareSG_bw_terra)

kde_childcareSG_bw_terra
```

Once we confirm that it is the SpatRaster class, then we can move on to next steps. If not troubleshoot again.

## Assigning projection systems

crs() of terra is used to assign the CRS information on kde_childcareSG_bw_terra layer.

```{r}
crs(kde_childcareSG_bw_terra) <- "EPSG:3414"
kde_childcareSG_bw_terra
```

## Plotting KDE map with tmap

Finally,the raster in cartographic quality map will be displayed using tmap package.

```{r}
tm_shape(kde_childcareSG_bw_terra) + 
  tm_raster(col.scale = 
              tm_scale_continuous(
                values = "viridis"),
            col.legend = tm_legend(
            title = "Density values",
            title.size = 0.7,
            text.size = 0.7,
            bg.color = "white",
            bg.alpha = 0.7,
            position = tm_pos_in(
              "right", "bottom"),
            frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1.0)
```


# First Order SPAA at planning subzone level

## Geospatial Data wrangling 
### Extract Study area 

The following target planning areas are extracted. 

```{r}
pg <- mpsz_cl %>%
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>%
  filter(PLN_AREA_N == "JURONG WEST")
```

::: callout-note
It is always a good practice to review the extracted areas.It is now done below 
:::

```{r}
par(mfrow=c(2,2))
plot(st_geometry(pg), main = "Punggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")

```

### Creating owin object 

Now the sf objects are convered to owin since its required by spatstat

```{r}
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)
```

### Combining point events and owin 

```{r}

childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]

```


Now rescaling is done to transform the unit measurement from metre to kilometre 

```{r}
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

Now we plot the 4 study areas to look and compare

```{r}
par(mfrow=c(2,2))
plot(unmark(childcare_pg_ppp.km), 
  main="Punggol")
plot(unmark(childcare_tm_ppp.km), 
  main="Tampines")
plot(unmark(childcare_ck_ppp.km), 
  main="Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), 
  main="Jurong West")
```

## Clark and Evans Test 
### Choa Chu Kang Planning area

clarkevans.test() of spatstat is used to performs Clark-Evans test of aggregation for childcare centre in Choa Chu Kang planning area.

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

### Tampines Planning area

clarkevans.test() of spatstat is used to performs Clark-Evans test of aggregation for childcare centre in Choa Chu Kang planning area.

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

## Computing KDE surfaces by Planning area

```{r}
par(mfrow=c(2,2))
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Jurong West")
```


# Example 1: Opening up a new Preschool location drilling into specific branches. 

Now in this case study, we look at the different preschool locations within the 4 planning areas : Woodlands, Bishan, Orchard, Choa Chu Kang. 

Make the assumption that we want to see the outlook of expanding and opening a new preschool in these areas :
- Woodlands
- Bishan
- Orchard
- Choa Chu Kang

```{r}
#Load & prepare data
mpsz <- st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")

mpsz_cl <- st_transform(mpsz, 3414)
preschool <- st_transform(preschool, 3414)

# Extract planning areas
wd <- mpsz_cl %>% filter(PLN_AREA_N == "WOODLANDS")
bs <- mpsz_cl %>% filter(PLN_AREA_N == "BISHAN")
or <- mpsz_cl %>% filter(PLN_AREA_N == "ORCHARD")
ck <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
```

```{r}
# Create owin objects
wd_owin <- as.owin(wd)
bs_owin <- as.owin(bs)
or_owin <- as.owin(or)
ck_owin <- as.owin(ck)

# Convert cycling paths to ppp and subset by owin
preschool_ppp <- as.ppp(preschool)
preschool_wd_ppp <- preschool_ppp[wd_owin]
preschool_bs_ppp <- preschool_ppp[bs_owin]
preschool_or_ppp <- preschool_ppp[or_owin]
preschool_ck_ppp <- preschool_ppp[ck_owin]

# Rescale to kilometers
preschool_wd_ppp.km <- rescale.ppp(preschool_wd_ppp, 1000, "km")
preschool_bs_ppp.km <- rescale.ppp(preschool_bs_ppp, 1000, "km")
preschool_or_ppp.km <- rescale.ppp(preschool_or_ppp, 1000, "km")
preschool_ck_ppp.km <- rescale.ppp(preschool_ck_ppp, 1000, "km")
```

```{r}
# Plot KDE surfaces by planning area
par(mfrow=c(2,2))
plot(density(preschool_wd_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Woodlands")
plot(density(preschool_bs_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Bishan")
plot(density(preschool_or_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Orchard")
plot(density(preschool_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
```

# Example 2: Plotting an overview to see density outlook (KDE map with tmap)

In this example, we assume that we are analysts, and a client pharmacy : Best Pharmacy wants us to create a visual to show the density view of all current Retail pharmacy locations, to sift out potential competitors in Singapore. 

```{r}
# Load & prepare data 
pharm <- st_read("data/geospatial/RetailpharmacylocationsGEOJSON.geojson")
mpsz <- st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
pharm_proj <- st_transform(pharm, 3414)
mpsz_cl <- st_transform(mpsz, 3414)
```


```{r}
# Create owin and ppp objects
sg_owin <- as.owin(mpsz_cl)
pharm_ppp <- as.ppp(pharm_proj)
pharmSG_ppp <- pharm_ppp[sg_owin]

# Rescale to kilometers
pharmSG_ppp_km <- rescale.ppp(pharmSG_ppp, 1000, "km")

# Compute KDE & Convert to raster
kde_pharm <- density(pharmSG_ppp_km, 
                     sigma = bw.diggle, 
                     edge = TRUE, 
                     kernel = "gaussian")

kde_pharm_terra <- rast(kde_pharm)
crs(kde_pharm_terra) <- "EPSG:3414"

```

```{r}
# Plot
tm_shape(kde_pharm_terra) + 
  tm_raster(col.scale = 
              tm_scale_continuous(
                values = "viridis"),
            col.legend = tm_legend(
              title = "Density values",
              title.size = 0.7,
              text.size = 0.7,
              bg.color = "white",
              bg.alpha = 0.7,
              position = tm_pos_in("right", "bottom"),
              frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1.0)
```

# Example 3: Exploring nearest Neighbour Analysis for Pharmacies 
Now with the plot given out, the client Best Pharmacy wants to the neighbour analysis to be conducted for more specific insights. 

```{r}
# Subset points to the study area
pharmSG_ppp <- pharm_ppp[sg_owin]

# Clark-Evans Test WITHOUT CSR 
cat(" Clark-Evans Test WITHOUT CSR \n")
clarkevans.test(pharmSG_ppp,
                correction = "none",
                clipregion = "sg_owin",
                alternative = c("clustered"))


# Clark-Evans Test WITH CSR 
cat("\n Clark-Evans Test WITH CSR (Monte Carlo) \n")
clarkevans.test(pharmSG_ppp,
                correction = "none",
                clipregion = "sg_owin",
                alternative = c("clustered"),
                method = "MonteCarlo",
                nsim = 999)
```
The Clark-Evans test statistic is R = 0.25045 <1 in both tests (the standard Z-test and the Monte Carlo simulation). 

The p-value is extremely small (p < 2.2e-16 for the Z-test, and p = 0.001 for the Monte Carlo test).

Since both p-values < 0.05, we reject the null hypothesis (H₀) that the retail pharmacies are randomly distributed. Hence, concluding there is strong statistical evidence that the distribution of retail pharmacies in Singapore is highly clustered.

The R-value of 0.25 indicates that the average distance between a pharmacy and its nearest neighbor is roughly only 25% of what we would expect if they were distributed randomly. This represents a very strong degree of spatial clustering.


Therefore, retail pharmacies in Singapore are strategically located in dense, highly concentrated clusters rather than being spread out across the country, possibly being at central areas near the MRT or main shopping malls for increased accessibility.

Recommendation : Avoid entering already saturated hotspots where cannibalization is likely. Instead, use the plotted out map in example 2 to identify the geographical gaps in coverage should expansion be the aim of analysis. If not, more resources should be focused on the hotspots to deal with competitors.  