---
title: "Hands-on Exercise 5B"
author: "Chong Jia Min Heidi"
format: html
date: "September 23 2026"
date-modified: "last-modified"
execute:
  echo: true 
  eval: true 
  warning: false 
  freeze: true
  message: false
---

# Overview

Local Measures of Spatial Autocorrelation (LMSA) focus on the relationships between each observation and its surroundings, rather than providing a single summary of these relationships across the map. In this sense, they are not summary statistics but scores that allow us to learn more about the spatial structure in our data. The general intuition behind the metrics however is similar to that of global ones. Some of them are even mathematically connected, where the global version can be decomposed into a collection of local ones. One such example are Local Indicators of Spatial Association (LISA). Beside LISA, Getis-Ord’s Gi-statistics will be introduce as an alternative LMSA statistics that present complementary information or allow us to obtain similar insights for geographically referenced data.

# Getting Started

## Importing of Data

These packages will be used (the same as per exercise 5A)

- **sf** is use for importing and handling geospatial data in R,

- **tidyverse** is mainly use for wrangling attribute data in R,

- **sfdep** will be used to compute spatial weights, global and local spatial autocorrelation statistics

- **tmap** will be used to prepare cartographic quality chropleth map.

As a note, knitr and KableExtra are packages used to create any tables for easier reading and understanding of knowledge if necessary.

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse,knitr,kableExtra)
```

We will be using the Hunan data from the In-class exercise 4 and Hands on 5A. The below will be done for importing the data

- import the Hunan shapefile and parse it into a sf polygon feature object

- Import Hunan_2012.csv file and parse it into a tibble data.frame

```{r}
# Hunan county boundary layer 
hunan <- st_read("data/geospatial/hunan/Hunan.shp") %>%
  st_transform(crs = 32650)

#This is the csv Hunan local development indicators  
#This will be in the  R dataframe class 
hunan_2012<- read_csv("data/aspatial/Hunan_2012.csv")
```

## Data joining

Now the relational join is conducted and the required data columns are retained (the unnecessary data columns are removed to not waste memory and space!)

```{r}
hunan <- left_join(hunan,hunan_2012) %>%  
  select(1:4, 7, 15)
```

## Data Plotting

Finally a basemap and a choropleth map showing the distribution of GDPPC 2012 by using `qtm()` of **tmap** package will be plotted out for initial reading.

```{r}

#| fig-width: 16
#| fig-height: 6
#| echo: false
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

::: callout-note
The plot reveals clear clusters on both maps. In the east Central region, the dark blue intensity shows that in this east central region there is a high GDP cluster, signalling these are the richest counties in Hunan, a secondary High-GDP area can be seen in the South of Hunan showing that another high GDP region to note in Hunan resides at South Hunan.

The other areas, particularly in the West shows a very light colour intensity. These form a low-value region across the Western Edge of hunan.

Both plots also indicate spatial clustering in the High GDP clusters East-Central region and Southeast Region, where the cold spots lie in western Hunan.
:::

# Local Indicators of Spatial Association (LISA)

Local Indicators of Spatial Association or LISA are statistics that evaluate the existence of clusters and/or outliers in the spatial arrangement of a given variable. For instance if we are studying distribution of GDP per capita of Hunan Provice, People Republic of China, local clusters in GDP per capita mean that there are counties that have higher or lower rates than is to be expected by chance alone; that is, the values occurring are above or below those of a random distribution in space.

## Computing Contiguity Spatial Weights

Before we can compute the local spatial autocorrelation statistics, we need to construct a spatial weights of the study area. The spatial weights is used to define the neighbourhood relationships between the geographical units (i.e. county) in the study area.

In the code chunk below, [`st_contiguity()`](https://sfdep.josiahparry.com/reference/st_contiguity) of **sfdep** package is used to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries. If you look at the documentation you will see that you can pass a “queen” argument that takes TRUE or FALSE as options. If you do not specify this argument the default is set to TRUE, that is, if you don’t specify queen = FALSE this function will return a list of first order neighbours using again the Queen criteria.

```{r}
wm_q <- hunan %>% mutate(nb = st_contiguity(geometry), wt = st_weights( nb, style = "W"), .before = 1)
```

```{r}
summary(wm_q$nb)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbour.

## Row-standardised Weights Matrix

Next, we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight (style=“W”). This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values.

While this is the most intuitive way to summaries the neighbors’ values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data.

For this example, we’ll stick with the style=“W” option for simplicity’s sake but note that other more robust options are available, notably style=“B”.

These steps if noticed is actually similar to that in Exercise 5A! :D

## Computing Local Moran's I

```{r}
set.seed(1234)
lisa <- wm_q %>% 
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(local_moran)
```

```{r}
glimpse(lisa)
```

## Mapping Local Moran's I values

```{r}
tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Morans'I",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Morans'I of GDPPC (Queen's method)")
```

### Mapping Local Moran I's P values

The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunks below produce a choropleth map of Moran’s I p-values by using functions of **tmap** package.

```{R}
tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")
```

### Combination of both plots

For effective interpretation, it is better to plot both the local Moran’s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Moran's I of GDPPC (Queen's method)")

p_ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("left", "bottom")
              )) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")

tmap_arrange(ii.map, p_ii.map, asp=1, ncol=2)
```

### Preparing & visualising LISA Map

Visualizing LISA involves creating maps that highlight significant spatial patterns, such as hot spots (high values clustered together) and cold spots (low values clustered together), as well as identifying spatial outliers. Common methods include choropleth maps showing LISA values, Moran scatterplots which depict local spatial autocorrelation for individual locations, and hot spot analysis maps.

The LISA Cluster Map shows the significant locations color coded by type of spatial autocorrelation. The first step before we can generate the LISA cluster map is to plot the Moran scatterplot.

#### Scatter plot

A Moran Scatterplot is a graphical tool in spatial analysis that visualizes the relationship between a variable’s values and the average values of its neighboring locations (its spatial lag) to identify spatial autocorrelation. By plotting standardized values of the variable on the x-axis and its spatial lag on the y-axis, the plot’s slope represents Moran’s I, revealing patterns of high-high or low-low value clusters and high-low or low-high value outliers.

In the code chunk below, [`st_lag()`](https://sfdep.josiahparry.com/reference/st_lag) of **sfdep** package is used to calculates the spatial lag of GDPPC given a neighbor (i.e. nb) and weights list (i.e. wt).

```{R}
lisa <- lisa %>%
  mutate(lag_GDPPC = st_lag(
    GDPPC, nb, wt),
    .before = 1) %>%
  unnest(lag_GDPPC)

ggplot(data = lisa, 
       aes(x = GDPPC, 
           y = lag_GDPPC)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "red") +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot") +
  theme_minimal()
```

To support effective visual communication, the code chunk below is used.

```{r}
ggplot(data = lisa, 
       aes(x = GDPPC, 
           y = lag_GDPPC, 
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "black") +
  geom_hline(yintercept=mean(lisa$lag_GDPPC), lty=2) + 
  geom_vline(xintercept=mean(lisa$GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red", 
               "Low-Low" = "blue",
               "Low-High" = "lightblue", 
               "High-Low" = "pink")) +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

Notice that the plot is split in 4 quadrants. The top right corner belongs to areas that have high GDPPC and are surrounded by other areas that have the average level of GDPPC. This are the high-high locations in the lesson slide.

#### Scatterplot with standardised variable

First we will use [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) to centers and scales the variable. Here centering is done by subtracting the mean (omitting NAs) the corresponding columns, and scaling is done by dividing the (centered) variable by their standard deviations.

```{r}
lisa <- lisa %>%
  mutate(z_GDPPC = scale(GDPPC),
         z_lag_GDPPC = scale(lag_GDPPC),
         .before = 1)
```

The [*as.vector()*](https://www.rdocumentation.org/packages/pbdDMAT/versions/0.5-1/topics/as.vector) added to the end is to make sure that the data type we get out of this is a vector, that map neatly into out dataframe.

Now, we are ready to plot the Moran scatterplot again by using the code chunk below.

```{r}
ggplot(data = lisa, 
       aes(x = z_GDPPC, 
           y = z_lag_GDPPC, 
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "black") +
  geom_hline(yintercept=mean(lisa$z_lag_GDPPC), lty=2) + 
  geom_vline(xintercept=mean(lisa$z_GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red", 
               "Low-Low" = "blue",
               "Low-High" = "lightblue", 
               "High-Low" = "pink")) +
  labs(x = "Standardised GDPPC",
       y = "Standardised Spatial Lag of GDPPC",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

### Preparing LISA Map Classes

The code chunks below show the steps to prepare a LISA cluster map.

Next, derives the spatially lagged variable of interest (i.e. GDPPC) and centers the spatially lagged variable around its mean.

This is follow by centering the local Moran’s around the mean.

Next, we will set a statistical significance level for the local Moran.

```{r}
signif <- 0.05       
```

### Plotting & Visualising LISA Map

By default, the LISA classes are stored in *mean*, *median* and *pysal* fields of lisa sf data frame. To prepare the LISA map, we will derive a new field called LISA-cluster from either one of these three field by using the code chunk below.

```{r}
lisa <- lisa %>%
  mutate(
    LISA_cluster = ifelse(
      p_ii < 0.05,
      as.character(mean), 
      "Insignificant"),
    LISA_cluster = factor(
      LISA_cluster,
      levels = c("Insignificant", "Low-Low", "Low-High", "High-Low", "High-High")
    )
  )
```

::: callout-note
Notice that a new class called *Insignificant* has been added for records whereby p_ii \> 0.05.
:::

Now we can plot the LISA map out.

```{R}
lisa_map <- tm_shape(lisa) + 
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "blue",        # Low-Low
        "lightblue",   # Low-High
        "pink",        # High-Low
        "red"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "LISA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("Local Moran's I Clusters (p < 0.05)")
lisa_map
```

For effective interpretation, it is better to plot both the local Moran’s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
#| label: fig-lisa-both
#| fig-width: 16
#| fig-height: 7
#| echo: false

tmap_arrange(ii.map, lisa_map, ncol = 2, asp = 1)
```

We can also include the local Moran’s I map and p-value map as shown below for easy comparison.

::: callout-note
The Local Moran's I map shows most counties in Hunan fall into the two lightest blue bins (0.0–1.9), indicating weakly positive local autocorrelation. A compact dark blue cluster appears in the east-central region (bins 2.0–5.0), while scattered pink/orange outliers (bins −2.0 to −0.1) appear in the west, southwest, and north.

The LISA cluster map shows most of Hunan is grey (Insignificant). Only two significant clusters emerge: a contiguous red High-High block (\~10–12 counties) in the east-central region, and a contiguous blue Low-Low block (\~8–10 counties) in the southwest. One or two light-blue Low-High counties sit north of the red cluster, and no pink High-Low counties appear.

Overall, significant clustering is confined to two compact regions, with the vast majority of counties statistically insignificant.
:::

# Hot & Cold Spots Analysis (HCSA)

Hot spots and cold spots analysis is a technique used in spatial statistics to identify statistically significant clusters of high values (hot spots) and low values (cold spots) of a geographically referenced attribute (i.e. GDPPC). These analyses help visualize and understand spatial patterns, revealing areas where data points are clustered more densely than would be expected by random chance.

Figure below shows the hot spots and cold spots of GDP per capita of Hunan Provice, People Republic of China.

The term hot spots refer to areas where high values are concentrated and surrounded by other high values. It has been used generically across disciplines to describe a region or value that is higher relative to its surroundings (Lepers et al 2005, Aben et al 2012, Isobe et al 2015). For example, a hot spot analysis might reveal clusters of high GDP per capita in a province, indicating areas with significantly higher GDP per capita than other parts of the province.

On the other hands, the term cold spots refer to areas where low values are concentrated and surrounded by other low values. An example could be clusters of low GDP per capita, indicating areas where GDP per capita is consistently low.

HCSA uses statistical methods, like the Getis-Ord Gi\* statistic, to determine if the observed clustering is statistically significant or could have occurred randomly.

The analysis consists of four steps:

- Deriving spatial weight matrix

- Computing Gi\* statistics

- Mapping Gi\* statistics

- Communicating the analysis results

## Deriving Distance Weights matrix

Whist the spatial autocorrelation considered units which shared borders, for Getis-Ord Gi\* statistics, we are defining neighbours based on distance.

There are two type of distance-based proximity matrix, they are:

- fixed distance weight matrix,

- adaptive distance weight matrix, and

- inverse distance weights.

### Computing fixed distance weights

Firstly, we need to determine the upper limit for distance band by using the steps below:

```{r}
ct <- critical_threshold(st_geometry(hunan))
ct
```

The summary report shows that the largest first nearest neighbour distance is 62.82km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

Using the critical threshold value computed above, code chunk below is used to compute the fixed distance weight.

```{r}
hunan_fdw <- hunan %>% 
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry), 
      upper = ct)),
    wt = st_weights(nb, 
                    style = "W"),
    .before = 1) 
```

### Computing Adaptive Distance weights

One of the characteristics of fixed distance weight matrix is that more densely settled areas (usually the urban areas) tend to have more neighbours and the less densely settled areas (usually the rural counties) tend to have lesser neighbours. Having many neighbours smooth the neighbour relationship across more neighbours.

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry as shown in the code chunk below.

```{R}
hunan_adw <- hunan %>%
  mutate(nb = include_self(
    st_knn(
      st_geometry(geometry),
      k = 6)),
    wt = st_weights(
      nb, style = "W"),
    .before = 1)
```

## Computing Local Gi\*

Local Gi\* (pronounced “G-i-star”) is a Getis-Ord statistic used in spatial analysis to identify statistically significant hot spots and cold spots of high or low attribute values within a dataset. It measures the spatial clustering of values by comparing the total value within a specified neighborhood to the global total, producing a Z-score and p-value for each location to determine the likelihood that the observed pattern is not due to random chance.

```{R}
HCSA_fdw <- hunan_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)

```

## Preparing and Visualising HCSA Map

```{r}
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 6,
                values = "brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "Gi*",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Gi* of GDPPC (Fixed Bandwidth d = 60799.91m)")
```

### Map local Gi Values

The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunk below plots a choropleth map of p-values of Gi\* by using functions of **tmap** package.

```{R}
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "simulated p-value",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (Fixed distance)")
```

### Plotting both visualisations together

```{r}
Gi_star_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Gi* of GDPPC")

p_values_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim", 
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("left", "bottom")
              )) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (fixed distance)")

tmap_arrange(Gi_star_map, p_values_map, asp=1, ncol=2)
```

### plot HCSA map

By default, the HCSA classes are stored in *cluster* field of *HCSA_fdw* sf data frame. To prepare the HSCA map, we will derive a new field called *HCSA_cluster* from *cluster* field by using the code chunk below.

```{r}
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low"  ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )
```

Now the HCSA plot is plotted

```{r}
HCSA_map <- tm_shape(HCSA_fdw) + 
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "red",        # Low-Low
        "blue"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "HSCA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("HCSA Clusters (p < 0.05)")
HCSA_map
```

For effective interpretation, it is better to plot both the local Gi\* values map and its corresponding HCSA map next to each other.

```{R}
tmap_arrange(Gi_star_map, HCSA_map, asp=1, ncol=2)
```

- A single small red Hot spot also appears isolated in the west, disconnected from the main eastern cluster. Overall, significant clustering is confined to a few compact regions, with the east-central Hot spot and the northwest Cold spot being the most prominent.

::: callout-note
The Local Gi\* map shows that most counties in Hunan fall into the light blue and pale pink bins, indicating weak local values.

A concentrated dark red zone is seen in the east-central region, surrounded by orange counties, showing a strong local peak. A secondary orange area is shown at the north of this core, while the west and southwest remain dominated by light blue.

The HCSA cluster map shows most of Hunan is grey (Insignificant). Only two types of significant clusters appear: a compact red Hot spot block (\~8–10 counties) in the east-central region, and three separate blue Cold spot blocks.

A single small red Hot spot also appears isolated in the west, disconnected from the main eastern cluster. Overall, significant clustering is confined to a few compact regions, with the east-central Hot spot and the northwest Cold spot being the most prominent.

- northwest

- west

- south-central area.

A single small red Hot spot also appears isolated in the west, disconnected from the main eastern cluster. Overall, significant clustering is confined to a few compact regions, with the east-central Hot spot and the northwest Cold spot being the most prominent.
:::
