---
title: "In class exercise 5"
---

# Getting started 

## loading in packages 

```{r}
pacman::p_load(sf,sfdep,tmap,tidyverse)
```

## Joining 

```{r}
# Hunan county boundary layer
hunan <- st_read("data/geospatial/hunan/Hunan.shp")

#This is the csv Hunan local development indicators 
#This will be in the  R dataframe class
hunan_2012<- read_csv("data/aspatial/Hunan_2012.csv")

hunan_GDPPC <- left_join(hunan,hunan_2012) %>%
  select(1:3, 7, 15,16,31,32)
```

# Global measures of spatial Association 

## 1. Deriving Queen Contiguity weights: sfdep methods 

```{r}
wm_q <- hunan_GDPPC %>% 
  mutate(nb=st_contiguity(geometry),
         wt=st_weights(nb,
                       style = "W"),
         .before=1)
```

## 2. Computing Global Moran'I

in the code chunk below, global_moran() function is used to compute the moran's I value. Different from spdep package, the output is a tibble data.frame.

```{r}
moranI<-global_moran(wm_q$GDPPC,
                     wm_q$nb,
                     wm_q$wt)
glimpse(moranI)
```

### performing global Moran permutation test 

In practice, Monte carlo simulation should be used to perform statistical test. For sfdep its supposed by global_moran_perm() function.

```{r}
set.seed(1234)

global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim=99)
```

::: callout-note
- The statistical report shows the p value is smaller than alpha value of 0.05. Hence we have enough statistical evidence to reject the null hypothesis that the spatial distribution of GDP per capita resembles a random distribution.

- Because Moran's I statistic is greater than 0 (positive spatial autocorrelation) it confirms a clustered pattern. This means similar values are near each other (High GDP per Capita countries are next to each other high GDP counties and low next to low.
:::

## 3. Computing Local Morans I

In this section you will learn how to compute local moran I of GDPPC at county level by using Local_moran() of sfdep package

```{r}
lisa<-wm_q %>%
  mutate(local_moran=local_moran(
    GDPPC,nb,wt,nsim=99),
    .before=1)%>%
  unnest(local_moran)
```

### plotting LISA map 

in the lisa sf data.frame we can find 3 fields containig the LISA categories : mean, median, pysal. Calssification in mean will be used as shown in code chunk below.

```{r}
lisa_sig<-lisa%>%
  filter(p_ii_sim<0.05)
tmap_mode("plot")
tm_shape(lisa)+
  tm_polygons()+
  tm_borders(fill_alpha =0.05) +
tm_shape(lisa_sig)+
  tm_polygons(fill="mean",
              fill.legend = tm_legend(
                text.size=1.15,
                width=6))+
  tm_borders(fill_alpha=0.4)+
  tm_layout(legend.title.size=1.25)

```
