---
title: "In-Class Exercise 4"
author: "Chong Jia Min Heidi"
format: html
date: "September 19 2026"
date-modified: "last-modified"
execute:
  echo: true 
  eval: true 
  warning: false 
  freeze: true
  message: false
---

# Getting started

## Loading packages

```{r}
pacman::p_load(sf, ggstatplot, tmap, tidyverse, knitr,GWmodel)
```

## Loading and Preparing Data

We will

- import the Hunan shapefile and parse it into a sf polygon feature object

- Import Hunan_2012.csv file and parse it into a tibble data.frame

- Join hunan and hunan_2012 data frames and tidy the data table by retaining NAME_2,ID_3, NAME_3, Country, GDPPC, GIO, Agri and Service

So step 1, import the data.

```{r}
# Hunan county boundary layer
hunan <- st_read("data/geospatial/hunan/Hunan.shp")

#This is the csv Hunan local development indicators 
#This will be in the  R dataframe class
hunan_2012<- read_csv("data/aspatial/Hunan_2012.csv")
```

Second, we will ensure that hunan is a polygon and the hunan_2012 csv is a tibble dataframe

```{r}
class(hunan) 
class(hunan_2012)
```

Finally, we will join the 2 datasets

```{r}
hunan_sf <- left_join(hunan,hunan_2012) %>%
  select(1:3, 7, 15,16,31,32)
```

::: callout-note
1.  Just do the joining to the 2 datasets first (without the select line)
2.  Go to R studio environment interface (top right) and pick out the columns we want
3.  Add the select line with the columns we want.
:::

# Mapping GDPPC

```{r}
# To edit the entire plot size, can experiment around with it :D
#| fig-width: 16
#| fig-height: 6
#| echo: false
basemap <- tm_shape(hunan_sf) + 
  tm_polygons()+
  tm_text("NAME_3",size=0.5)

gdppc <- qtm(hunan_sf,"GDPPC")+
  tm_layout(
    legend.position = c("left","bottom"))
tmap_arrange(basemap,gdppc,asp=1,ncol=2)
```

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

# Geographically Weighted Summary Statistics (GSWW) with adaptive bandwidth

## Determine adaptive Bandwidth

### cross validation

```{r}

bw_CV <- bw.gwr(GDPPC ~ 1,# only picking one variable which is GDPpercapita
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = TRUE)
```

### AICc

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
                  data = hunan_sp,
                  approach = "AICc",
                  adaptive = TRUE,
                  kernel = "bisquare",
                  longlat = TRUE)

```

### Computing geographically weighted summary statistics

The code chunk below computes geographically weighted summary statistics by using gwss() of GWmodel package. The analysis is performed using adaptive distance weighted. The distance is derived using AICc method and the kernel method is bi-square.

```{r}
gwstat<-gwss(data=hunan_sp,
             vars="GDPPC",
             bw=bw_AIC,
             kernel="bisquare",
             quantile=TRUE, 
             adaptive=TRUE,
             longlat = T)
```

::: callout-note
gwss support 5 kernel methods , namely Gaussian, exponential, bi-square, Tricube and boxcar. Repeat the code chunk above with different kernel methods and compare their results visually.
:::

## Preparing Output data

::: callout-important
gwss() with quantile argument computes eight local summary statistics, namely: local means,local standard deviations, local variance, local skew, local coefficients of variation, local medians, local interquartile ranges, and local quantile imbalances. they are output in SDF data frame.
:::

Code chunk below is used to extract SDF data table from gwss object output from gwss() for subsequent geovilisation. It will be converted into data.frame by using as.data.frame().

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)

```

Next the cbind() is used to append the newly derived data.frame onto *hunan_sf* sf data.frame

```{r}
hunan_gstat <-cbind(hunan_sf,gwstat_df)
```

# Visualising Geographially Weighted summary Statistics

```{r}
#| fig-width: 14
#| fig-height: 8
#| echo: false
tm_shape(hunan_gstat) + 
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) + 
  tm_borders(fill_alpha = 0.5) + 
  tm_title("Distribution of geographically weighted mean") + 
  tm_layout(legend.position = c("left", "bottom"))
```
