---
title: "Take-Home Exercise 1"
date: "September 12 2026"
date-modified: "last-modified"
---

# Overview

Public safety refers to the protection of the general public, communities, and critical infrastructure from crimes, disasters, accidents, and other major threats through coordinated government and agency action.

Geospatial analytics, supported by multi-source geospatial data and geospatial data science and analysis methods, provides powerful means to transform location-based data into visual evidence, analytical insights, and decision support. It plays an important role in areas such as law enforcement, emergency management, disaster response, transportation safety, and environmental management.

A spatial point pattern represents a collection of geographic locations at which particular events occur, such as forest-fire detections or road traffic accidents. A spatio-temporal point pattern extends this representation by incorporating the time at which each event occurs, allowing analysts to investigate how the distribution and interaction of events vary across both space and time.

Understanding where and when public-safety events occur can help planners and managers identify areas of concern, understand patterns of event occurrence, allocate resources, and develop evidence-informed strategies for prevention and response.

## Objectives

This take-home exercise investigates the spatial and/or spatio-temporal patterns of public-safety point events using one the case study of **Road traffic accidents** within the Thailand with emphasis on the Bangkok Metropolitan Region (BMR) for the selected study year 2022.

In Thailand, it is known that they have recorded one of the [highest road death tolls in the world](https://www.nationthailand.com/in-focus/40010768), which emphasizes the importance of having this analysis within the exercise to gain proper insights to help make better decisions and actionable(s) to reduce the accident rates in the country and in lieu the death toll.

The specific objectives from this exercise are :

- To prepare and integrate real-world geospatial and temporal data for spatial and spatio-temporal point-pattern analysis.

- To investigate first-order properties of the point pattern by examining how the intensity or distribution of events varies across space and/or time.

- To investigate second-order properties by examining whether events exhibit evidence of spatial or spatio-temporal interaction, such as clustering, inhibition, or approximate randomness.

- To critically evaluate the analytical assumptions, methodological choices, and limitations associated with the selected point-pattern methods.

- To communicate the analytical findings effectively through appropriate geovisualisation and geocommunication.

- To translate the analytical evidence into meaningful implications for public-safety planning and management.

# Data

These are the datasets used for the exercise

## Thailand Road Accident Dataset

<https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022>

This is a comprehensive dataset on recorded road accidents in Thailand, spanning from approximately 2019 to 2022. For the sake of the exercise only the latest data of 2022 will be used here.

::: callout-note
The data is from the Office of the Permanent Secretary, Ministry of Transport, where for easier visualisation the [public dashboard](https://trams.mot.go.th/main-dashboard) can be accessed as well.
:::

## Thailand Boundary Map

<https://www.geoboundaries.org/countryDownloads.html>

This is the Thailand Boundary map published by the Royal Thai Survey Department, OCHA ROAP on the [geoBoundaries](https://www.geoboundaries.org/) website in 2019 (This is also the latest version available for public download).

# Task 1: Getting Started - Preparation of Data

## Importing of data

First the required libraries are installed/loaded into file.

```{r}
pacman::p_load(sf, spatstat, sparr, stpp, tmap, tidyverse, lubridate,gt)
```

Now we load in both datasets.

```{r}
# Thailand Road Accident Dataset
acc<- read_csv("data/aspatial/thai_road_accident_2019_2022.csv")
```

Here we will check the summary of the **Thailand Road Accident** Dataset.

```{r}
head(acc) 
```

Now we will load in the second dataset : Thailand Boundary Map.

This dataset

```{r}
# Thailand Boundary Map
thai_map <- st_read("data/geospatial/Thai/geoBoundaries-THA-ADM2.shp")

# Column names
names(thai_map)

# Preview attributes without geometry
st_drop_geometry(thai_map) %>% head(10)
st_drop_geometry(thai_map) %>% glimpse()

# How many unique values per column
st_drop_geometry(thai_map) %>% summarise(across(everything(), n_distinct))

```

## Data Inspection & Cleaning.

A Point event is basically defined as a single record within the dataset, or a specific road traffic accident Each event will be geolocated using the `longitude` and `latitude` columns.

So it is important that any duplicated rows of data do not exist and we handle all missing or invalid coordinates, which will be done in this sub-section.

Several things will checked upon within this subsection. These include

- Handling of Invalid/Missing Locations

- Removal of Unnecessary Columns & rows

- Ensuring Unique Rows

**Missing and Invalid Coordinates**\
Rows with missing or invalid coordinates (NA in `latitude` or `longitude`) were identified and handled. First, we checked for rows with unknown or NA provinces. Because the latitude and longitude columns provide specific point locations but lack the contextual information needed for readable insights (e.g., "which province?"), we prioritized recovering the province attribute. We successfully recovered 25 out of 34 missing provinces (73.5%) by spatially joining the coordinates with the ADM1 shapefile. The remaining 9 rows, which lacked both coordinates and province data, were excluded from the analysis.

### Handling Invalid/Missing Locations

First, to ensure there is no invalid rows (error due to location column issues, we will check for rows with unknown or NA provinces).

This is important since during analysis when writing out the insights, we will link back to the province column rather than the latitude and longitutde column for readers to understand better. The latitude and longitude gives the specific point of the accident but with just these 2 columns readers dont know which city or province we are going to talk about.

::: callout-note
If there are any unknown province locations, check for unknown latitude and longitude. To see how much can be recovered since Lat and Long gives the specific point. If we have latitude and longitude in the missing columns, it is possible to recover province.
:::

```{r}

unknown_rows <- acc %>%
  filter(province_en == "unknown" | is.na(province_en))

unknown_with_xy <- unknown_rows %>%
  filter(!is.na(latitude), !is.na(longitude))

unknown_no_xy <- unknown_rows %>%
  filter(is.na(latitude) | is.na(longitude))


tibble(
  category      = c("Unknown province (total)",
                    "→ with coordinates (recoverable)",
                    "→ without coordinates (need manual or drop)"),
  n             = c(nrow(unknown_rows),
                    nrow(unknown_with_xy),
                    nrow(unknown_no_xy)),
  pct_of_unknown = round(100 * n / nrow(unknown_rows), 1)
)
```

So we can see that there are 25 coordinates that can actually be recovered from the 34 with missing province locations. This is good since 73.5% of missing values can be recovered.

Through analysis of the shapefile, ADM 2 looks into the district and not province. So it will not be possible to use our current ADM 2 shapefile of Thailand to attain the province values to need to link the coordinates. We will use ADM 1 instead. Now the data will be loaded in.

```{r}

provinces <- st_read("data/geospatial/Thai_province/geoBoundaries-THA-ADM1.shp") %>%
  st_transform(4326)

# Check the province name column to ensure we can use ADM1 
names(provinces)


# Isolate the rows we want to recover, so it dosent affect the main dataset
unknown_with_xy <- acc %>%
  filter(
    (province_en == "unknown" | is.na(province_en)) &
      !is.na(latitude) & !is.na(longitude)
  )

pts <- st_as_sf(
  unknown_with_xy,
  coords = c("longitude", "latitude"),
  crs = 4326
)

# Here is where the join is done to connect the datapoints
pts_joined <- st_join(pts, provinces, join = st_within)
pts_joined %>%
  st_drop_geometry() %>%
  select(acc_code, shapeName) %>%
  head()
missed <- pts_joined %>% filter(is.na(shapeName))
nrow(missed)

# In case of any misses with the provinces, this is a fallback to link up to nearest province 
if (nrow(missed) > 0) {
  idx <- st_nearest_feature(missed, provinces)
  pts_joined$shapeName[is.na(pts_joined$shapeName)] <- provinces$shapeName[idx]
}

# Merge accounts back to actual dataset
acc_joined <- acc %>%
  left_join(
    pts_joined %>%
      st_drop_geometry() %>%
      select(acc_code, recovered_province = shapeName),
    by = "acc_code"
  ) %>%
  mutate(
    province_en = ifelse(
      province_en == "unknown" | is.na(province_en),
      recovered_province,
      province_en
    )
  ) %>%
  select(-recovered_province)

acc <- acc_joined
```

```{r}
# Verification, now there should be 9 missing data points rather than 34!
acc %>%
  filter(is.na(province_en)) %>%
  nrow()
```

Now that we've settled the 24 recoverable records of data, we will now remove the 9 missing data points where it is impossible to recover.

```{r}
acc<- acc %>%
  filter(!is.na(province_en))

# Verify that there are now 0 rows with missing province
acc %>%
  filter(is.na(province_en)) %>%
  nrow()
```

### Removal of Unnecessary Rows

The current dataset is from 2019-2022, since we are aiming to only look at 2022 data we want to drop the unnecessary rows (year 2019-2021).

::: callout-note
We will only look at the range of year 2022 through the use of incident_datetime since we want to analyse the accidents that occurred in 2022 instead of report_datetime (refers to the date and time when the accident is recorded)

This is important because there time is needed after the accident has occurred for it to be recorded, resulting in a reporting lag. Therefore, this distinction is highlighted to show that the correct column is used for the filtering together with the importance of this distinction.
:::

```{r}
# The date and Time will be parsed to Thailand timezone, so hardware (computer) which is in Singapore timezone can accurately filter the data to year 2022. 
acc <- acc %>%
  mutate(
    incident_datetime = ymd_hms(incident_datetime, tz = "Asia/Bangkok"),
    report_datetime   = ymd_hms(report_datetime,   tz = "Asia/Bangkok")
  )


# Filter so that only year 2022 data exists in current dataset 
acc<- acc %>%
  filter(year(incident_datetime) == 2022)

# Verify that the datetime is what we need
nrow(acc)
range(acc$incident_datetime)  

```

Now below is a summary of the rows removed up till now.

```{r}

tibble(
  Stage = c(
    "Raw import",
    "After province recovery + drop unrecoverable",
    "After temporal filter (year 2022)"
  ),
  Rows = c(81735, 81701, 21006),
  Removed = c(NA, 34, 60695)
) %>%
  gt() %>%
  tab_header(title = "Row Counts at Each Cleaning Stage") %>%
  cols_label(Stage = "Cleaning Stage", Rows = "Rows Remaining", Removed = "Rows Removed") %>%
  fmt_number(columns = c(Rows, Removed), decimals = 0)
```

### Removal of Columns

Now we look at Removal of columns, where province_th and route columns will be removed.

::: callout-note
From the previous code input, we can see the first few rows of the data is displayed to have a rough sense of the data.

There is a "Duplicated" Column where "province_th" is the same as "province_en" but in the Thai language. Since we cannot understand Thai, we will remove this column instead of province_en.

Furthermore, route column is redundant since the coordinates are given through the latitude and longitude data columns. The inconsistent description of junctions, highway numbers and sectors are also the reason why this column should not be used, since it will only cause more confusion in analysis process.
:::

```{r}
# Drop unncessary columns 
acc<- acc %>% select(-route, -province_th)

# Now to confirm the correct columns are dropped. We will expect to see 16 columns instead. (6x16)
head(acc) 

```

### Ensuring Unique rows in data

Ephasizing again on the importance of having each point to represent an independent event, it is imperative that every accident within the dataset is unique!

Therefore, to make sure there is no duplicated rows `any(duplicated(acc))` is used to ensure that every accident (full row) is unique.

To double check, we also use the `any(duplicated(acc))` to verify that there is no duplicate or repeated observations in the cleaned dataset which looks at the account code itself which is said to be unique. Especially since any(duplicated(acc)) will not be able to detect that there is duplicate record of E.g Accident ABC if any other column within the 2 rows are different.

```{r}
# This looks for duplicates across the entire row, where it only records if all values within one row is the same as another
any(duplicated(acc))

# This ensures that all acccount codes (acc_code) is different and unique, so that we arent looking at multiple accident accounts. 
sum(duplicated(acc$acc_code))
```

Since FALSE and 0 is returned its confirmed all records are unique, and we can proceed on with drawing out the plots.

## Assumptions & Decisions made

### Assumptions made

Some assumptions made in this exercise for the analysis to be valid. The list is as follows

#### Reliability of Data

There is always the possibility of human error, where in this case human error can refer to personnel forgetting to record an accident that has occurred. So we will assume here that the data provided is fully representative of all accidents that occurred in 2022.

#### Homogeneity

It is assumed that the risk of every accident occuring across Thailand is uniform, or exactly the same. In reality, there are many variables that can influence this risk level like the volume of traffic or level of maintenance of the roads or etc. This assumption will be put to allow exploratory trend detection.

#### Independence

We assume that each accident is independent of each other. There can be times in reality where an accident can occur due to the occurrence of another, this is known as multi-vehicle crashes. Hence multiple points in this case will be recorded within these datasets, but we will assume in this exercise that all accidents are recorded as separate events.

#### Isotropy

We assume here that the intensity of the accidents is the same in all directions. In reality, road networks are linear, so accidents are constrained to roads. So with the assumption, the analysis can be a relatively useful form of visualisation on risk.

### Decisions made

The **Coordinate Reference System (CRS)** was decided to be changed within the Thailand Boundary from WGS 84 (EPSG:4326) to UTM Zone 47N (EPSG:32647). This is essential because UTM coordinates are in meters, which allows distance-based methods (bandwidth selection, Ripley's K, edge correction) to operate on meaningful planar distances rather than degrees.

## Overview plot

This overview plot is made to ensure that the data is loaded properly and there are no errors in data manipulation or cleaning.

```{r}
acc <- acc %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

acc <- st_transform(acc, 32647)
thai_map <- st_transform(thai_map,    32647)


```

```{r}
# Now we plot an overview on the tmap. 
tm_shape(thai_map) +
  tm_polygons(col = "grey95", border.col = "grey70", lwd = 0.1) +
  tm_shape(acc) +
  tm_dots(col = "red", size = 0.02, alpha = 0.5) +
  tm_layout(frame = FALSE)
```

::: callout-note
The reason why Tmap is being used instead of the usual plot() :

tmap is a package helps to render the ploygons better than plots, preventing overlapping borders like what plot() will do. Basically it is better at rendering and plotting the multi-ploygon background more clearly.
:::

# Task 2: First-Order Point Pattern Analysis

## Computing accident STKDE by Month

What we currently have is only the overall accident distribution in the entire year of 2022, but we cannot see where and when the accidents would occur together.

With the STKDE by month helping to estimate a 3D intensity surface on 1 plot per month for a total of 12 plots, it gives a smoothed continuous surface to help compare the intensity of accidents. We can see which areas within Thailand have a high accident intensity and in which months the intensity is concentrated.

Therefore STKDE by month is done for the 1st Order Point Analysis because it provides the insights on the where and when the accident happens together, or the analysis of both spatial and temporal dimensions simultaneously.

### Step 1: Converting to OWIN format

Our current file is in the sf object with 70 polygon regions. However, for the `ppp` and STKDE to be conducted we need to transform it to a single OWIN format to conduct the First-Order Pattern Analysis.

```{r}
# Union all provinces into a single polygon
thailand_sf <- provinces %>%
  st_make_valid() %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32647)

# Convert to OWIN
thailand_owin <- as.owin(thailand_sf)

# We put in class and summary to ensure that our we have indeed converted this to OWIN format as required. 
class(thailand_owin)
summary(thailand_owin)
```

Since current class here is now OWIN we are good to go to step 2.

### Step 2 : Extracting Data by Month

when we use `ppp` , it only accepts numeric or character marks. However currently we only have the date (incident_datetime) by a date format. Therefore, we will have to extract out the `Month_num` (1 to 12 from January to December) and `Month_fac` to produce readable facet labels without inline conversion.

`DayofYear` is also extracted for now in case of the need for analysis of each specific day of the year.

```{r}

acc <- acc %>%
  mutate(
    DayofYear = yday(incident_datetime),
    Month_num = month(incident_datetime),
    Month_fac = month(incident_datetime, label = TRUE, abbr = FALSE)
  )

```

### Step 3 : Visualising accidents by month

Before computing the KDE, it is important to plot a faceted point map (one small map of Thailand per month).

This is a first order visualisation of the data to understand the spatial distribution of accidents across the year and if there is any seasonality before computing the KDE. It also acts as the baseline before the STKDE is done to allow a comparison to be done.

```{r}
tm_shape(thailand_sf) +
  tm_polygons() +
  tm_shape(acc) +
  tm_dots(size = 0.1) +
  tm_facets(by = "Month_fac",
            free.coords = FALSE,
            drop.units = TRUE)

```

::: callout-important
**Analysis of plot.**

From the plot, it can be seen that April and December have a heavier distribution throughout the months in the year, which ties in the monthly trends in Thailand with higher vehicle numbers on the road from the [Songkran Holiday travelling in April](https://he02.tci-thaijo.org/index.php/agstjournal/article/view/275930) and New Year Holiday travels in December.

However it is to note that in all 12 months, there is a darker concentration around the Bangkok and central provinces proving that Bangkok area is a permanent high-risk accident zone. In contrast, the North-Eastern Region in Thailand is showing fewer dots in the visuals throughout the year reflecting a lower accident volume assuming that the data provided is fully representative of all accidents that occurred in 2022 (as mentioned in assumptions made).
:::

### Step 4 : coverting to ppp

Since we are looking at the Month specifically we will keep only the Month number column and use it to convert to ppp.

It is also recommended to check again here to see if there are duplicated units. Usually this is done to prevent any inaccuracies when analysing data. However in this case, it is expected for there to be multiple crashes at the same intersection so it's fine for there to be duplicate values. The presence of duplicate values if any violates the strict independence assumption of point pattern analysis, but the events are retained because they represent genuinely distinct accidents. Therefore, the duplicate check is just to note that there are these sort of multi-crash accidents, and highlight the presence of this analytical limitation when conducting the analysis.

```{r}

# Keep only Month_num column
acc_month <- acc %>% select(Month_num)

# Convert to ppp
acc_month_ppp <- as.ppp(acc_month)

# Check for duplicates 
any(duplicated(acc_month_ppp))

```

### Step 5 : Compute STKDE

Now the STKDE is finally computed.

```{r}
acc_month_owin <- acc_month_ppp[thailand_owin]
summary(acc_month_owin)
st_kde <- spattemp.density(acc_month_owin)
summary(st_kde)

```

The `spattemp.density()` function has estimated a Spatial bandwidth of 38159.91 (\~38.16km). This is with the smoothing distance in UTM Zone 47N metres. It is known that a larger `h` produces a smoother, more generalised surface while a smaller `h` reveals local hot spots. Overall it is decided that the spatial bandwidth should not be further changed it balances detail against noise given the irregular shape of Thailand's boundary.

The temporal bandwidth is estimated to be lambda = 0.0079 (in units of the temporal range). Since time was rescaled to `[1, 12]` months, we have a narrow temporal smoothing window of about 0.1 month, which means that the density surface will change sharply month-to-month which is best to emphasize the high intensity time scope of April and December. Therefore edge correction is applied to reduce the bias near Thailand's border, where points outside the boundary cannot be observed but still contribute to the intensity estimate.

he trivariate density is normalised and conditioned to allow each monthly slice to integrate to 1 and make for direct visual comparison across months.

### Step 6 : Plot out STKDE

After the computation, the STKDE is plotted out.

```{r}

tims <- c(1,2,3,4,5,6,7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

::: callout-important
**Analysis of plot.**

The STKDE visualizations confirm the initial findings from our raw monthly map. More specifically :

1.  April and December peaks in accident numbers with January coming behind as a close third.

2.  Bangkok regions is a hot spot for accident occurrence through the red colour intensity in those months

However it also refines a few points. First, the South peninsula area actually shows a low density per unit area rather than the entire area being a hot spot. The prominence of the entire South Peninsula was rather shown in the initial monthly plots because of the raw accident counts inflating the concentration of accident rates.

Furthermore, elaborating on intensity, other than Bangkok, the entire country shows a uniformly low accident intensity given the blue colour scale across all these areas in all months. This proves that the Bangkok Region is the single main dominant accident hot spot in Thailand for the year 2022.
:::

## Computing Death Toll using STKDE by Month

There are different levels of severity of vehicle accidents, from a light collision without injuries to heavy injuries in the ICU or even death.

Therefore, we will also look at the death toll within Thailand from accidents using STKDE by month again.

```{r}
# Filter to fatal accidents only
acc_fatal <- acc %>%
  filter(number_of_fatalities > 0) %>%
  mutate(
    Month_num = month(incident_datetime),
    Month_fac = month(incident_datetime, label = TRUE, abbr = FALSE)
  )

# Convert to ppp
acc_fatal_month <- acc_fatal %>% select(Month_num)
acc_fatal_ppp <- as.ppp(acc_fatal_month)
acc_fatal_owin <- acc_fatal_ppp[thailand_owin]

# Conduct STKDE
st_kde_fatal <- spattemp.density(acc_fatal_owin)
summary(st_kde_fatal)

# Plotting visualisation

par(mfcol = c(2, 3))
for (i in 1:12) {
  plot(st_kde_fatal, i, override.par = FALSE, fix.range = TRUE,
       main = paste("Fatal KDE at month", i))
}
```

::: callout-important
**Analysis of plot.**

Now this plot on the death toll reveals major different trends than the overall accident plots, suggesting that accident frequency and severity are following different spatial and temporal patterns.

Here, the fatal accident intensity is elevated in Northern, North-Eastern Thailand, and Bangkok in mainly the January and December months, where January has replaced April as the peak month in regards to fatality of accidents.

Since the rural roads are mainly in the North and North-East regions of Thailand, it fits the general fact that rural roads generates fewer but higher severity accidents leading to higher fatal accident rates and urban areas will have more low speed collision which is usually not as severe.
:::

## First Order SPAA by Highlighted Regions (Clark Evans) 

Now that the overview analysis of the entire country of Thailand is done, there are 3 regions that are flagged out to be accident hotspots or have a higher fatality intensity :

- **Bangkok Metropolitan Region (BMR)** — the dominant urban accident hotspot.
- **Northern Thailand** — rural, mountainous, elevated fatality intensity.
- **North-Eastern Thailand (Isan)** — rural, elevated fatality intensity.

We want to study these regions in more detail. Especially since the SPKDE only helps to say where in Thailand are the accidents more intense, and not if the specific accidents are clustered or dispersed. The Clark-Evans Test however, provides statistical evidence that the visual hot spot you saw in the STKDE is real, not a coincidence of plotting, which is also why we are looking at the highlighed Regions throuh the Clark Evans Testing for our First order SPAA.

The Clark-Evans Test assumes there is no duplicated points. So it is required to ensure that we check there is no duplicated points since any duplicated points will inflate the clustering signal.

### Step 1 : Defining the Study Regions

We have our regions that we want to drill into to study. However the regions are made up of a numerous number of provinces, where our data only has the column of provinces or districts. Therefore, we have to put this additional step to define what our study regions are.

This is a code to confirm on the provinces within our dataset as well as the spelling in each one of them before proceeding with further analysis. If the spelling is wrong, we will have issues in writing out the right code.

```{r}
#| label: define-regions
provinces %>%
  st_drop_geometry() %>%
  pull(shapeName) %>%
  sort() %>%
  print()
```

Therefore, we will be using the following provinces (decided after research on the definition of the Bangkok Metropolitan Region (BMR), Northern Thailand and North Eastern Thailand). As shown in the code chunk below, the 3 regions that we want to screen in for the analysis are now generated here.

```{r}
# BMR Region 
bmr_provinces <- c(
  "Bangkok",
  "Nonthaburi Province",
  "Pathum Thani Province",
  "Samut Prakan Province",
  "Samut Sakhon Province",
  "Nakhon Pathom Province"
)

# Northern Thailand
north_provinces <- c(
  "Chiang Mai Province",
  "Chiang Rai Province",
  "Lampang Province",
  "Lamphun Province",
  "Mae Hong Son Province",
  "Nan Province",
  "Phayao Province",
  "Phrae Province",
  "Uttaradit Province"
)

# North-Eastern Thailand
northeast_provinces <- c(
  "Kalasin",
  "Khon Kaen Province",
  "Chaiyaphum Province",
  "Nakhon Phanom Province",
  "Nakhon Ratchasima Province",
  "Bueng Kan Province",
  "Maha Sarakham Province",
  "Mukdahan Province",
  "Yasothon Province",
  "Roi Et Province",
  "Loei Province",
  "Sakon Nakhon Province",
  "Surin Province",
  "Nong Khai Province",
  "Nong Bua Lam Phu Province",
  "Amnat Charoen Province",
  "Udon Thani Province",
  "Ubon Ratchathani Province",
  "Buri Ram Province",
  "Si Sa Ket Province"
)
```

### Step 2 : Building the Region Objects

Now we will build the region objects

```{r}
make_region <- function(prov_names, region_name) {
  sf_obj <- provinces %>%
    filter(shapeName %in% prov_names) %>%
    st_make_valid() %>%
    st_transform(32647)
  owin_obj <- as.owin(sf_obj %>% st_union() %>% st_zm(drop = TRUE, what = "ZM"))
  cat(region_name, ":", nrow(sf_obj), "provinces\n")
  list(sf = sf_obj, owin = owin_obj)
}

bmr   <- make_region(bmr_provinces,       "BMR")
north <- make_region(north_provinces,     "Northern")
ne    <- make_region(northeast_provinces, "North-Eastern")
```

### Step 3 : Generating Clark-Evans and KDE

```{r}
# BMR
acc_bmr <- acc %>% st_filter(bmr$sf, .predicate = st_within)
bmr_ppp <- as.ppp(acc_bmr); Window(bmr_ppp) <- bmr$owin
ce_bmr  <- clarkevans.test(bmr_ppp, correction = "none")
ce_bmr

# Northern
acc_north <- acc %>% st_filter(north$sf, .predicate = st_within)
north_ppp <- as.ppp(acc_north); Window(north_ppp) <- north$owin
ce_north  <- clarkevans.test(north_ppp, correction = "none")
ce_north

# North-Eastern
acc_ne <- acc %>% st_filter(ne$sf, .predicate = st_within)
ne_ppp <- as.ppp(acc_ne); Window(ne_ppp) <- ne$owin
ce_ne  <- clarkevans.test(ne_ppp, correction = "none")
ce_ne
```

```{r}
par(mfrow = c(1, 3))

plot(
  st_geometry(bmr$sf),
  main = "Bangkok Metropolitan Region"
)

plot(
  st_geometry(north$sf),
  main = "Northern Thailand"
)

plot(
  st_geometry(ne$sf),
  main = "North-Eastern Thailand"
)
```

### Step 4: Creating OWIN objects

```{r}
bmr_owin   <- bmr$owin
north_owin <- north$owin
ne_owin    <- ne$owin

```

### Step 5: Combination of OWIN and Point events 

We have to put the acc

```{r}
acc_ppp <- as.ppp(acc)
class(acc_ppp)

```

```{r}

acc_bmr_ppp   <- acc_ppp[bmr_owin]
acc_north_ppp <- acc_ppp[north_owin]
acc_ne_ppp    <- acc_ppp[ne_owin]
```

### Step 6: Rescaling from m to km 

Currently the dataset is set to be in metres (UTM coordinates). So all the bandwidths and distances will be in enormous numbers within all the plots and calculations will be tough to read and interpret if the rescaling not be done.

```{r}
acc_bmr_ppp.km   <- rescale.ppp(acc_bmr_ppp,   1000, "km")
acc_north_ppp.km <- rescale.ppp(acc_north_ppp, 1000, "km")
acc_ne_ppp.km    <- rescale.ppp(acc_ne_ppp,    1000, "km")
```

### Step 7 : Plotting 

```{R}
par(mfrow = c(2, 2))

plot(unmark(acc_bmr_ppp.km), main = "BMR",cex = 0.5)
plot(unmark(acc_north_ppp.km), main = "Northern Thailand",cex = 0.5)
plot(unmark(acc_ne_ppp.km), main = "North-Eastern Thailand", cex = 0.5)

par(mfrow = c(1, 1))
```

note : The analysis of the plots are included within the Clark Evans statistical testing.

### Clark-Evans Statistical Testing 

Now that the plots are out and we can see where in the 3 regions are the accidents are clustered or spread out at, we conduct the Clark Evans statisitical testing to see if the results are significant.

#### BMR 

```{r}
clarkevans.test(acc_bmr_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)

sum(duplicated(acc_bmr_ppp))
```

Since p \< 2.2 × 10⁻¹⁶ \<0.05 it proves that the results are highly significant and the accidents in the Bangkok Metropolitan Region has a highly significant spatial clustering. With the R =0.265, it shows the observed average nearest neighbour distance is at 73.5% closer than normal to consider it as spatial randomness. Therefore with reference from the Clark Evans where there are very distinct lines on the highways of the Bangkok Metropolitan Region, these spots are where the concentrated accidents are which are the areas proven where the government needs to intervene to have the most impact in regards to the accident occurrences.

#### North-Eastern Thailand 

```{r}
clarkevans.test(acc_ne_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)

sum(duplicated(acc_ne_ppp))
```

Since p \< 2.2 × 10⁻¹⁶ \<0.05, it proves that the results are highly significant and the accidents in the Thailand Northern East Region has a highly significant spatial clustering. With the R =0.41935, it shows the observed average nearest neighbour distance is at 59.1% closer than normal to consider it as spatial randomness. Therefore with reference from the Clark Evans plot where there are darker intensity points plotted across several different areas in Northern East Thailand (where accident occurence area is more spread out across the region) , these spots are where the concentrated accidents are which are the areas proven where the government needs to intervene to have the most impact in regards to the accident occurrences.

#### Northern Thailand 

```{r}
clarkevans.test(acc_north_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)

sum(duplicated(acc_north_ppp))
```

Since p \< 2.2 × 10⁻¹⁶ \<0.05, it proves that the results are highly significant and the accidents in the Bangkok Metropolitan Region has a highly significant spatial clustering. With the R =0.3014, it shows the observed average nearest neighbour distance is at 70.9% closer than normal to consider it as spatial randomness. Therefore with reference from the Clark Evans plot where there are darker intensity points plotted in the bottom area of Northern Thailand, these spots are where the concentrated accidents are which are the areas proven where the government needs to intervene to have the most impact in regards to the accident occurrences.

## Analysis & Recommendations

From the analysis, it is known that Bangkok is a persistent hotspot for accidents to occur, especially in the holidays of April and December. However, in January and December the accident fatality intensity spikes at the North and Northeast Regions of Thailand.

It is proven that these observations are strongly significant given the p \< 2.2 × 10⁻¹⁶ \<0.05 in the Clark Evans test in all tested regions where accident intensity are shown to be stronger. And that all highlighted regions (BMR, Northern East, North Thailand) does have a highly significant spatial clustering.

Therefore, it is recommended for these measures to take place based on the insights :

### Accident Frequency

**Short term:** It would be imperative to implement more speed checks, DUI checkpoints and traffic patrols at the densely plotted out points within the 3 regions at the Clarks Evans plots especially in the months of January, April and December. This is to counter the spike in accident numbers due to the increased vehicle numbers on the road due to their main holiday periods. It will be a temporary measure to allow time to plan and deploy the long term solutions.

**Long term :** If the accident distribution in the Bangkok area is persistently high within Thailand then it suggests that regardless of the reason whether it is due to high traffic volume, poorly maintained roads or lack of traffic safety enforcement a targeted investigation on the root problem is required. After which interventions like road re-designing or traffic law enforcement are necessary to ensure safety in the future.

### Accident Severity

**Short term :** There will be a need for the Transport authority within Thailand to go on high alert and prioritize the emergency-response ability within the North and North Eastern provinces/districts during January and December, with more emphasis in the Northern Eastern Thailand due to the fact that the accident areas within NE Thailand is less clustered and more spread out thus increasing the risks. This will be to allow more efficient and faster emergency responses to the fewer but more severe (with higher risk of death) accidents, instead of focusing on high volume enforcement in Bangkok.

**Long Term :** Suggestion would be to plan, improve and implement better speed-limit enforcement infrastructure, Road-surface maintenance, and roadside safety measures like having more barriers, lighting, signages to help on visibility and safety of the drivers on the road in the rural areas.

# Task 3: Second-Order Point Pattern Analysis

Now, since we had identified that the Bangkok Metropolitan Region, North Eastern Region and North Thailand as the hot-spots for accidents to occur in Thailand, we will screen in to analyse more deeply within this region to get the most impact out of the analysis in preventing future accidents on the road.

For this Second Order Point Pattern Analysis, we will be utilising the K and L Functions. The reason is as stated below.

Clark Evans only looks at a one distance to the nearest point, which computes R and returns a p value for significance at a very short range scale (nearest neighbour distance). However, the K Function and L Function helps to provide a better idea on how clustering varies with distance which the previous Clark-Evans test does not provide.

Previously we learnt in Thailand Is there clustering at all? Now as a research question we want to understand at which distance does this clustering occur now that we had proven the existance of the clustering at these 3 main regions : Bangkok Metropolitan Region, North Eastern Region and North Thailand.

## K - Function 

The K-Function specifically measures the number of events found up to a given distance of any particular event.

### Assumptions made 

### Limitations/Caveats 

### Bangkok Metropolitan Region

First, the set seed is added in to ensure that the points stay the same in every calculation for easier analysis. Then we use the rescaled ppp variables to be put into the Ripley correction to get the best values before we do out the K Function plot for the Bangkok Metropolitan Region.

```{r}
set.seed(1234)
#acc_bmr_small <- acc_bmr_ppp.km[sample(npoints(acc_bmr_ppp.km), 1000)]
#K_bmr <- Kest(acc_bmr_small, correction = "Ripley")
```

```{r}
#plot(K_bmr, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

::: callout-note
Analysis :

tbw
:::

Now we will have to conduct the Complete Spatial Randomness Test to confirm the observed spatial patterns above. To do this, a hypothesis test will be conducted, The hypothesis and test are as follows:

Ho = The distribution of accidents at the Bangkok Metropolitan Region are randomly distributed.

H1= The distribution of accidents at the Bangkok Metropolitan Region are not randomly distributed.

```{r}
#K_bmr.csr <- envelope(acc_bmr_ppp.km, Kest, 
#                     nsim = 99, rank = 1,
#                     global=FALSE)
```

::: callout-note
Analysis :

tbw
:::

Now we conduct the final plotting

```{r}
#plot(K_bmr.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

::: callout-note
Analysis :

tbw
:::

### North Thailand 

First, the set seed is added in to ensure that the points stay the same in every calculation for easier analysis. Then we use the rescaled ppp variables to be put into the Ripley correction to get the best values before we do out the K Function plot for the Bangkok Metropolitan Region.

```{r}
set.seed(1234)
#acc_north_small <- acc_north_ppp.km[sample(npoints(acc_north_ppp.km), 1000)]
#K_north <- Kest(acc_north_small, correction = "Ripley")
```

```{r}
#plot(K_north, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

::: callout-note
Analysis :

TBW
:::

Now we will have to conduct the Complete Spatial Randomness Test to confirm the observed spatial patterns above. To do this, a hypothesis test will be conducted, The hypothesis and test are as follows:

Ho = The distribution of accidents at the Northern Thailand Region are randomly distributed.

H1= The distribution of accidents at the Northern Thailand Region are not randomly distributed.

```{r}
#K_north.csr <- envelope(acc_north_ppp.km, Kest, 
#                      nsim = 99, rank = 1,
#                      global=FALSE)
```

::: callout-note
Analysis :

TBW
:::

Now we conduct the final plotting

```{r}
#plot(K_north.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

::: callout-note
Analysis :

TBW
:::

### Northern East Thailand 

First, the set seed is added in to ensure that the points stay the same in every calculation for easier analysis. Then we use the rescaled ppp variables to be put into the Ripley correction to get the best values before we do out the K Function plot for the Northern East Region.

```{r}
#set.seed(1234)
#acc_ne_small <- acc_ne_ppp.km[sample(npoints(acc_ne_ppp.km), 1000)]
#K_ne <- Kest(acc_ne_small, correction = "Ripley")
```

```{r}
#plot(K_ne, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

::: callout-note
Analysis :

TBW
:::

Now we will have to conduct the Complete Spatial Randomness Test to confirm the observed spatial patterns above. To do this, a hypothesis test will be conducted, The hypothesis and test are as follows:

Ho = The distribution of accidents at the Northern East Region are randomly distributed.

H1= The distribution of accidents at the Northern East Region are not randomly distributed.

```{r}
#K_ne.csr <- envelope(acc_ne_ppp.km, Kest, 
#                        nsim = 99, rank = 1,
#                        global=FALSE)
```

::: callout-note
Analysis:

TBW
:::

Now we conduct the final plotting

```{r}
#plot(K_ne.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

::: callout-note
Analysis :

TBW
:::

# Task 4 : Integration & Interpretation 

Your discussion should address:

- What does the first-order analysis reveal about event intensity or distribution?

- What does the second-order analysis reveal about interaction among events?

- How do the first-order and second-order findings complement each other?

- Are the findings consistent across different analytical views?

- What assumptions or limitations should be considered when interpreting the results?

- Are there alternative explanations for the observed patterns?

Where appropriate, conduct a sensitivity or robustness analysis to determine whether important conclusions are sensitive to analytical choices such as bandwidth, distance range, temporal aggregation, or other parameters.

# Task 5 : Geovisualisation and Analytical Interpretation

Prepare appropriate geovisualisations and statistical graphics to communicate the findings.

For each major geovisualisation, provide a concise analytical interpretation of no more than 150 words.

The interpretation should explain:

- what the visualisation reveals;

- what evidence supports the interpretation; and

- any important limitations or caveats.

The emphasis should be on analytical clarity and effective communication, rather than decorative or aesthetic complexity.

# Task 6 : Public Safety Planning & Management 

With reference to the analytical findings, discuss how the results could support public-safety planning and management.

Your discussion should be no more than 500 words.

Depending on the selected case study, relevant implications may include:

- land-use planning;

- transportation and road-safety planning;

- emergency response;

- infrastructure planning;

- environmental management;

- resource allocation;

- risk mitigation; and

- public-safety policy

Your discussion should distinguish between evidence supported directly by the analysis and interpretations or recommendations that require additional information.

Avoid making causal claims unless they are supported by appropriate evidence.
