---
title: "Take-Home Exercise 1"
date: "September 12 2026"
date-modified: "last-modified"
---

# Overview

Public safety refers to the protection of the general public, communities, and critical infrastructure from crimes, disasters, accidents, and other major threats through coordinated government and agency action.

Geospatial analytics, supported by multi-source geospatial data and geospatial data science and analysis methods, provides powerful means to transform location-based data into visual evidence, analytical insights, and decision support. It plays an important role in areas such as law enforcement, emergency management, disaster response, transportation safety, and environmental management.

A spatial point pattern represents a collection of geographic locations at which particular events occur, such as forest-fire detections or road traffic accidents. A spatio-temporal point pattern extends this representation by incorporating the time at which each event occurs, allowing analysts to investigate how the distribution and interaction of events vary across both space and time.

Understanding where and when public-safety events occur can help planners and managers identify areas of concern, understand patterns of event occurrence, allocate resources, and develop evidence-informed strategies for prevention and response.

## Objectives

This take-home exercise investigates the spatial and/or spatio-temporal patterns of public-safety point events using one the case study of **Road traffic accidents** within the Bangkok Greater Metropolitan Area for the selected study year 2022.

The specific objectives from this exercise are :

- To prepare and integrate real-world geospatial and temporal data for spatial and spatio-temporal point-pattern analysis.

- To investigate first-order properties of the point pattern by examining how the intensity or distribution of events varies across space and/or time.

- To investigate second-order properties by examining whether events exhibit evidence of spatial or spatio-temporal interaction, such as clustering, inhibition, or approximate randomness.

- To critically evaluate the analytical assumptions, methodological choices, and limitations associated with the selected point-pattern methods.

- To communicate the analytical findings effectively through appropriate geovisualisation and geocommunication.

- To translate the analytical evidence into meaningful implications for public-safety planning and management.

# Data

These are the datasets used for the exercise

## Thailand Road Accident Dataset

<https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022>

This is a comprehensive dataset on recorded road accidents in Thailand, spanning from approximately 2019 to 2022. For the sake of the exercise only the latest data of 2022 will be used here.

::: callout-note
The data is from the Office of the Permanent Secretary, Ministry of Transport, where for easier visualisation the [public dashboard](https://trams.mot.go.th/main-dashboard) can be accessed as well.
:::

## Thailand Boundary Map

<https://www.geoboundaries.org/countryDownloads.html>

This is the Thailand Boundary map published by the Royal Thai Survey Department, OCHA ROAP on the [geoBoundaries](https://www.geoboundaries.org/) website in 2019 (This is also the latest version available for public download).

# Getting Started - Preparation of Data

## Importing of data

First the required libraries are installed/loaded into file.

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

Now we load in both datasets.

```{r}
# Thailand Road Accident Dataset
acc<- read_csv("data/aspatial/thai_road_accident_2019_2022.csv")
```

Here we will check the summary of the **Thailand Road Accident** Dataset.

```{r}
head(acc) 
```

Now we will load in the second dataset : Thailand Boundary Map

```{r}
# Thailand Boundary Map
map <- st_read("data/geospatial/Thai/geoBoundaries-THA-ADM2.shp")

# Column names
names(map)

# Preview attributes without geometry
st_drop_geometry(map) %>% head(10)
st_drop_geometry(map) %>% glimpse()

# How many unique values per column
st_drop_geometry(map) %>% summarise(across(everything(), n_distinct))

```

## Data Inspection & Cleaning.

A Point event is basically defined as a single record within the dataset, or a specific road traffic accident Each event will be geolocated using the `longitude` and `latitude` columns.

So it is important that any duplicated rows of data do not exist and we handle all missing or invalid coordinates, which will be done in this sub-section.

Several things will checked upon within this subsection. These include

- Handling of Invalid/Missing Locations

- Removal of Unnecessary Columns & rows

- Ensuring Unique Rows

**Missing and Invalid Coordinates**\
Rows with missing or invalid coordinates (NA in `latitude` or `longitude`) were identified and handled. First, we checked for rows with unknown or NA provinces. Because the latitude and longitude columns provide specific point locations but lack the contextual information needed for readable insights (e.g., "which province?"), we prioritized recovering the province attribute. We successfully recovered 25 out of 34 missing provinces (73.5%) by spatially joining the coordinates with the ADM1 shapefile. The remaining 9 rows, which lacked both coordinates and province data, were excluded from the analysis.

### Handling Invalid/Missing Locations 

First, to ensure there is no invalid rows (error due to location column issues, we will check for rows with unknown or NA provinces).

This is important since during analysis when writing out the insights, we will link back to the province column rather than the latitude and longitutde column for readers to understand better. The latitude and longitude gives the specific point of the accident but with just these 2 columns readers dont know which city or province we are going to talk about.

::: callout-note
If there are any unknown province locations, check for unknown latitude and longitude. To see how much can be recovered since Lat and Long gives the specific point. If we have latitude and longitude in the missing columns, it is possible to recover province.
:::

```{r}

unknown_rows <- acc %>%
  filter(province_en == "unknown" | is.na(province_en))

unknown_with_xy <- unknown_rows %>%
  filter(!is.na(latitude), !is.na(longitude))

unknown_no_xy <- unknown_rows %>%
  filter(is.na(latitude) | is.na(longitude))


tibble(
  category      = c("Unknown province (total)",
                    "→ with coordinates (recoverable)",
                    "→ without coordinates (need manual or drop)"),
  n             = c(nrow(unknown_rows),
                    nrow(unknown_with_xy),
                    nrow(unknown_no_xy)),
  pct_of_unknown = round(100 * n / nrow(unknown_rows), 1)
)
```

So we can see that there are 25 coordinates that can actually be recovered from the 34 with missing province locations. This is good since 73.5% of missing values can be recovered.

Through analysis of the shapefile, ADM 2 looks into the district and not province. So it will not be possible to use our current ADM 2 shapefile of Thailand to attain the province values to need to link the coordinates. We will use ADM 1 instead. Now the data will be loaded in.

```{r}

provinces <- st_read("data/geospatial/Thai_province/geoBoundaries-THA-ADM1.shp") %>%
  st_transform(4326)

# Check the province name column to ensure we can use ADM1 
names(provinces)


# Isolate the rows we want to recover, so it dosent affect the main dataset
unknown_with_xy <- acc %>%
  filter(
    (province_en == "unknown" | is.na(province_en)) &
      !is.na(latitude) & !is.na(longitude)
  )

pts <- st_as_sf(
  unknown_with_xy,
  coords = c("longitude", "latitude"),
  crs = 4326
)

# Here is where the join is done to connect the datapoints
pts_joined <- st_join(pts, provinces, join = st_within)
pts_joined %>%
  st_drop_geometry() %>%
  select(acc_code, shapeName) %>%
  head()
missed <- pts_joined %>% filter(is.na(shapeName))
nrow(missed)

# In case of any misses with the provinces, this is a fallback to link up to nearest province 
if (nrow(missed) > 0) {
  idx <- st_nearest_feature(missed, provinces)
  pts_joined$shapeName[is.na(pts_joined$shapeName)] <- provinces$shapeName[idx]
}

# Merge accounts back to actual dataset
acc_joined <- acc %>%
  left_join(
    pts_joined %>%
      st_drop_geometry() %>%
      select(acc_code, recovered_province = shapeName),
    by = "acc_code"
  ) %>%
  mutate(
    province_en = ifelse(
      province_en == "unknown" | is.na(province_en),
      recovered_province,
      province_en
    )
  ) %>%
  select(-recovered_province)

acc <- acc_joined
```

```{r}
# Verification, now there should be 9 missing data points rather than 34!
acc %>%
  filter(is.na(province_en)) %>%
  nrow()
```

Now that we've settled the 24 recoverable records of data, we will now remove the 9 missing data points where it is impossible to recover.

```{r}
acc<- acc %>%
  filter(!is.na(province_en))

# Verify that there are now 0 rows with missing province
acc %>%
  filter(is.na(province_en)) %>%
  nrow()
```

### Removal of Unnecessary Rows 

The current dataset is from 2019-2022, since we are aiming to only look at 2022 data we want to drop the unnecessary rows (year 2019-2021).

::: callout-note
We will only look at the range of year 2022 through the use of incident_datetime since we want to analyse the accidents that occurred in 2022 instead of report_datetime (refers to the date and time when the accident is recorded)

This is important because there time is needed after the accident has occurred for it to be recorded, resulting in a reporting lag. Therefore, this distinction is highlighted to show that the correct column is used for the filtering together with the importance of this distinction.
:::

```{r}
# The date and Time will be parsed to Thailand timezone, so hardware (computer) which is in Singapore timezone can accurately filter the data to year 2022. 
acc <- acc %>%
  mutate(
    incident_datetime = ymd_hms(incident_datetime, tz = "Asia/Bangkok"),
    report_datetime   = ymd_hms(report_datetime,   tz = "Asia/Bangkok")
  )


# Filter so that only year 2022 data exists in current dataset 
acc<- acc %>%
  filter(year(incident_datetime) == 2022)

# Verify that the datetime is what we need
nrow(acc)
range(acc$incident_datetime)  

```

### Removal of Columns

Now we look at Removal of columns, where province_th and route columns will be removed.

::: callout-note
From the previous code input, we can see the first few rows of the data is displayed to have a rough sense of the data.

There is a "Duplicated" Column where "province_th" is the same as "province_en" but in the Thai language. Since we cannot understand Thai, we will remove this column instead of province_en.

Furthermore, route column is redundant since the coordinates are given through the latitude and longitude data columns. The inconsistent description of junctions, highway numbers and sectors are also the reason why this column should not be used, since it will only cause more confusion in analysis process.
:::

```{r}
# Drop unncessary columns 
acc<- acc %>% select(-route, -province_th)

# Now to confirm the correct columns are dropped. We will expect to see 16 columns instead. (6x16)
head(acc) 

```

### Ensuring Unique rows in data 

Ephasizing again on the importance of having each point to represent an independent event, it is imperative that every accident within the dataset is unique!

Therefore, to make sure there is no duplicated rows `any(duplicated(acc))` is used to ensure that every accident (full row) is unique.

To double check, we also use the `any(duplicated(acc))` to verify that there is no duplicate or repeated observations in the cleaned dataset which looks at the account code itself which is said to be unique. Especially since any(duplicated(acc)) will not be able to detect that there is duplicate record of E.g Accident ABC if any other column within the 2 rows are different.

```{r}
# This looks for duplicates across the entire row, where it only records if all values within one row is the same as another
any(duplicated(acc))

# This ensures that all acccount codes (acc_code) is different and unique, so that we arent looking at multiple accident accounts. 
sum(duplicated(acc$acc_code))
```

Since FALSE and 0 is returned its confirmed all records are unique, and we can proceed on with drawing out the plots.

## Assumptions & Decisions made 

### Assumptions made 

Some assumptions made in this exercise for the analysis to be valid. The list is as follows

#### Homogeneity

It is assumed that the risk of every accident occuring across Thailand is uniform, or exactly the same. In reality, there are many variables that can influence this risk level like the volume of traffic or level of maintenance of the roads or etc. This assumption will be put to allow exploratory trend detection.

#### Independence 

We assume that each accident is independent of each other. There can be times in reality where an accident can occur due to the occurrence of another, this is known as multi-vehicle crashes. Hence multiple points in this case will be recorded within these datasets, but we will assume in this exercise that all accidents are recorded as separate events.

#### Isotropy

We assume here that the intensity of the accidents is the same in all directions. In reality, road networks are linear, so accidents are constrained to roads. So with the assumption, the analysis can be a relatively useful form of visualisation on risk.

### Decisions made 

The **Coordinate Reference System (CRS)** was decided to be changed within the Thailand Boundary from WGS 84 (EPSG:4326) to UTM Zone 47N (EPSG:32647). This is essential because UTM coordinates are in meters, which allows distance-based methods (bandwidth selection, Ripley's K, edge correction) to operate on meaningful planar distances rather than degrees.

## Overview plot

This overview plot is made to ensure that the data is loaded properly and there are no errors in data manipulation or cleaning.

```{r}
acc <- acc %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

acc <- st_transform(acc, 32647)
map <- st_transform(map,    32647)
```

```{r}
# Now we plot an overview on the tmap. 
tm_shape(map) +
  tm_polygons(col = "grey95", border.col = "grey70", lwd = 0.1) +
  tm_shape(acc) +
  tm_dots(col = "red", size = 0.02, alpha = 0.5) +
  tm_layout(frame = FALSE)
```

::: callout-note
The reason why Tmap is being used instead of the usual plot() :

tmap is a package helps to render the ploygons better than plots, preventing overlapping borders like what plot() will do. Basically it is better at rendering and plotting the multi-ploygon background more clearly.
:::

# First-Order Point Pattern Analysis

## Computing STKDE by Month 

Why do I want to compute STKDE by month? What do i expect to be able to analyse by doing this?

```{r}

library(sf)
library(spatstat)

# 1. Union all 928 district polygons into a SINGLE polygon
#    st_union() dissolves internal boundaries
thailand_union <- map %>%
  st_union() %>%
  st_sf()   # wrap back into an sf object for as.owin()

# 2. Convert the single-polygon sf to an owin
thailand_owin <- as.owin(thailand_union)

# 3. Plot to verify the shape looks like Thailand
plot(thailand_owin, main = "Thailand Study Window (owin)")



# Create the Month_num column in acc_sf
acc$Month_num <- as.numeric(format(acc$incident_datetime, "%m"))
acc_month <- acc %>%
  select(Month_num)

# Now we convert to ppp
acc_month_ppp <- as.ppp(acc_month)

# set window to thailand boundary 
Window(acc_month_ppp) <- thailand_owin


# Plot to confirm the window and points look right
plot(acc_month_ppp, main = "Accidents with Month Marks")


```

```{r}
# Now we print out a summary to ensure the class is correct
acc_month_ppp
summary(acc_month_ppp)
```

It is also recommended to check again here to ensure there are no duplicated units which can

1.  Cause data loading to be a lot slower

2.  Cause inaccuracies when analysing data

If output is FALSE, we are good to go!

```{r}
any(duplicated(acc_month_ppp))
```

Now, the following code will combine *acc_am_ppp* and *map_owin* objects into one. To also ensure the correctness/accuracy of the overall output object,we will plot out the result of the combination.

Conduct appropriate first-order spatial and/or spatio-temporal point-pattern analyses using the prepared data.

Your analysis should investigate how the intensity or distribution of events varies across space and/or time.

You should:

- select appropriate analytical methods; justify why the selected methods are appropriate;

- explain important analytical parameters and assumptions;

- produce appropriate statistical graphics and geovisualisations; and

- interpret the resulting patterns in the context of your selected public-safety problem.
